Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-06-29 workflow/workflow.Rmd
Updates to text.

---
title: "Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, UC Berkeley
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine
  - name: Kelly N. Street
    affilitation: Graduate Group in Biostatistics, UC Berkeley
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, UC Berkeley
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics, UC Berkeley, and Department of Statistics, UC Berkeley

@@

keywords: single-cell, single-cell genomics, RNA-seq, gene expression, zero inflation, normalization, clustering, lineage inference, statistical models, workflow, R, Bioconductor

@@

#bioconductor:
library(clusterExperiment)
library(BiocParallel)
library(zinbwave)
#github
library(slingshot)
#library(scone)
#CRAN
library(doParallel)
library(RColorBrewer)
library(gam)
set.seed(20)
if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
```

@@

# count matrix
E <- assayData(Cufflinks_eSet)$counts_table

@@

```{r}
# remove ERCC and CreER genes
cre <- E["CreER",]
ercc <- E[grep("^ERCC-", rownames(E)),]
E <- E[grep("^ERCC-", rownames(E), invert = TRUE), ]
E <- E[-which(rownames(E)=="CreER"), ]
dim(E)
```

@@

# symbol for cell not assigned to a lineage in original data
metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2

@@

```{r scone, fig.cap="Plot generated by scone to filter low quality cells.", eval=FALSE}
# Metric-based Filtering sample-filtering
data("housekeeping")
hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]

@@

```{r,eval=FALSE}
# Filtering to top 1000 most variable
vars <- rowVars(log1p(assay(se)))
names(vars) <- rownames(se)
vars <- sort(vars, decreasing = TRUE)
core <- se[names(vars)[1:1000],]
```

@@

```{r rsec_50}
fn <- '../data/RSEC_W_combineMinSize10.rda'
if (runClus & !file.exists(fn)){
  #symbol for samples missing from original clustering
  seObj <- SummarizedExperiment(t(W), colData = colData(core))
  print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
                                  betas = 0.8, dimReduce="none",
                clusterFunction = "hierarchical01", minSizes=1,
                ncores = NCORES, isCount=FALSE,
                subsampleArgs = list(resamp.num=100,
                                     clusterFunction="kmeans",
                                     clusterArgs=list(nstart=10)),
                seqArgs = list(k.min=3, top.can=5), verbose=TRUE,
                combineProportion = 0.7,
                mergeMethod = "none", random.seed=424242,
                combineMinSize = 10)))
  save(ceObj, file = fn)
}else{
  load(fn)
}
```

@@

## Differential expression analysis

>>>>

---
title: "Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine
  - name: Kelly N. Street
    affilitation: Graduate Group in Biostatistics, University of California, Berkeley
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley

@@

keywords: single-cell, single-cell genomics, RNA-seq, gene expression, zero inflation, normalization, dimensionality reduction, clustering, lineage inference, differential expression, workflow, R, Bioconductor

@@

# Bioconductor:
library(clusterExperiment)
library(BiocParallel)
library(zinbwave)
# GitHub
library(slingshot)
#library(scone)
# CRAN
library(doParallel)
library(RColorBrewer)
library(gam)
set.seed(20)
if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
```

@@

# Count matrix
E <- assayData(Cufflinks_eSet)$counts_table

@@

```{r}
# Remove ERCC and CreER genes
cre <- E["CreER",]
ercc <- E[grep("^ERCC-", rownames(E)),]
E <- E[grep("^ERCC-", rownames(E), invert = TRUE), ]
E <- E[-which(rownames(E)=="CreER"), ]
dim(E)
```

@@

# Symbol for cells not assigned to a lineage in original data
metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2

@@

```{r scone, fig.cap="Plot generated by scone to filter low quality cells.", eval=FALSE}
# QC-metric-based sample-filtering
data("housekeeping")
hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]

@@

```{r,eval=FALSE}
# Filtering to top 1000 most variable genes
vars <- rowVars(log1p(assay(se)))
names(vars) <- rownames(se)
vars <- sort(vars, decreasing = TRUE)
core <- se[names(vars)[1:1000],]
```

@@

```{r rsec_50}
fn <- '../data/RSEC_W_combineMinSize10.rda'
if (runClus & !file.exists(fn)){
  # Symbol for samples missing from original clustering
  seObj <- SummarizedExperiment(t(W), colData = colData(core))
  print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
                                  betas = 0.8, dimReduce="none",
                clusterFunction = "hierarchical01", minSizes=1,
                ncores = NCORES, isCount=FALSE,
                subsampleArgs = list(resamp.num=100,
                                     clusterFunction="kmeans",
                                     clusterArgs=list(nstart=10)),
                seqArgs = list(k.min=3, top.can=5), verbose=TRUE,
                combineProportion = 0.7,
                mergeMethod = "none", random.seed=424242,
                combineMinSize = 10)))
  save(ceObj, file = fn)
}else{
  load(fn)
}
```

@@

## Differential expression analysis along lineages
